{RandomRationalPoints} package for \emph{Macaulay2}

### 2020-12-22 FastLinAlg/Paper/RandomRationalPoints.tex
Format changes

\documentclass[11pt]{amsart}
\usepackage{calc,amssymb,amsthm,amsmath,fullpage    }
%\usepackage{mathtools}
\RequirePackage[dvipsnames,usenames, table]{xcolor}
\usepackage{hyperref}
\usepackage[capitalize]{cleveref}
\hypersetup{
bookmarks,
bookmarksdepth=3,
bookmarksopen,
bookmarksnumbered,
pdfstartview=FitH,
colorlinks,backref,hyperindex,
linkcolor=Sepia,
anchorcolor=BurntOrange,
citecolor=MidnightBlue,
citecolor=OliveGreen,
filecolor=BlueViolet,
menucolor=Yellow,
urlcolor=OliveGreen
}
\usepackage{xypic}
\usepackage{alltt}
\usepackage{multicol}  
%\usepackage{etex}
\usepackage{xspace}
\usepackage{rotating}
\interfootnotelinepenalty=100000
\theoremstyle{definition}
\newtheorem{example}{Example}[section]

%\usepackage{mabliautoref}
%\usepackage{colonequals}
%\frenchspacing
%\input{kmacros3.sty}
%\usepackage{stmaryrd}
%
\usepackage{verbatim}
\usepackage{enumerate}
\begin{document}
\title{{RandomRationalPoints} package for \emph{Macaulay2}}
\author{Sankhaneel Bisui}
\email{sbisui@tulane.edu}
\address{Tulane University}
\author{Zhan Jiang}
\email{zoeng@umich.edu}
\address{University of Michigan}
\author{Sarasij Maitra}
\email{sm3vg@virginia.edu}
\address{University of Virginia}
\author{Thai Nguyen}
\email{tnguyen11@tulane.edu}
\address{Tulane University}
\author{Karl Schwede}
\email{schwede@math.utah.edu}
\address{Department of Mathematics, University of Utah, 155 S 1400 E Room 233, Salt Lake City, UT, 84112}
\date{\today}

@@

\keywords{RandomRationalPoints, Macaulay2}
%\thanks{Martinova was supported by a University of Utah Mathematics REU fellowship and by the University of Utah ACCESS program.  
%%Yao was supported by a University of Utah Mathematics REU fellowship.}

@@

%We start {\tt randomCoordinateChange}, a helper function that also might be useful to others.

%	\item {\tt Codimension}: It takes input an integer. Various functions need to know the codimension/height of the scheme/ideal it is working with.  Setting this to be an integer will tell the function not to compute the codimension and to use this value instead.  The default value is {\tt null}, in which case the function will compute the codimension.  
%%%
%%Codimension is not a function, it is an option.
%%%
\subsection*{\tt randomCoordinateChange:} This function takes a polynomial ring as an input and outputs the coordinate change map.  
	i.e. given a polynomial ring, this will produce a linear automorphism of the ring.  This function checks whether the map is an isomorphism by computing the Jacobian.

@@

	{{\small\color{blue}
	\begin{verbatim}
	i3 : R = ZZ/11[x,y,z];

@@

	i5 : matrix randomCoordinateChange(R, MaxCoordinatesToReplace=>1)

@@

	i6 : matrix randomCoordinateChange(R, MaxCoordinatesToReplace=>1, Homogeneous=>false)

@@

	i7 : matrix randomCoordinateChange(R, MaxCoordinatesToReplace=>1, Replacement=>Binomial)

@@

%\begin{enumerate}
\subsection{genericProjection} 
This function finds a random (somewhat, depending on options) generic projection of the ring or ideal.
The typical usages are as follows: 
\vspace{1em}

@@

i1 : R=ZZ/5[x,y,z,w];

i2 : I = ideal(x,y^2,w^3+x^2);

i3 : genericProjection(2,I)

            ZZ                                         2          2
o6 = (map(R,--[z, w],{- x - 2y - z, - y - 2z}), ideal(z  - z*w - w ))                   
            5

o6 : Sequence
	\end{verbatim}
	}}
\end{example}

@@

i1 : R=ZZ/5[x,y,z,w];

i2 : I = ideal(x,y^2,w^3+x^2);

i3 : genericProjection(2,R/I)

@@

o3 : Sequence
	\end{verbatim}
	}}
\end{example}

@@

i2 : R=ZZ/5[x,y,z,w];

i3 : I = ideal(x,y^2,w^3+x^2);

@@

i4 :genericProjection(2,I, Replacement=>Binomial)

@@

o4 : Sequence
\end{verbatim}
}}
\end{example}

%\vspace{1em} 
\subsection{\tt projectionToHypersurface} This function creates a projection to a hypersurface. The typical usages are as follows:

@@

\vspace{1em}
\begin{example}
	{{\small\color{blue}
	\begin{verbatim}
	i2 : R=ZZ/5[x,y,z];

	i3 : I = ideal(random(3,R)-2, random(2,R))

@@

	\end{verbatim}
}}
\end{example}

%\end{enumerate}
\section{randomPoints}\label{randomPoints}
{\tt randomPoints} is a function to find rational or geometric points in a variety. The typical usages are as follows:

@@

	Notice that the speed, or success, varies depending on the strategy.
\newpage

@@

			i2 : R = ZZ/101[x,y,z];

			i3 : J = ideal(x^3+y^2+1,z^3-x^2-y^2+2);

@@

	\end{verbatim}
	}}
	\end{example}

@@

		{{\small\color{blue}
		\begin{verbatim}
	i1: R = ZZ/101[x,y,z];

	o1 : PolynomialRing

	i2 : J = ideal "x3+y2+1,z3-x2-y2+2";

@@

	i3 : randomPoints(J,Strategy=>BruteForce,PointCheckAttempts=>20000)
		-- used 0.84679 seconds

@@

	o3 : List
	\end{verbatim}
	}}
	\end{example}

@@

	When calling {\tt randomPoints}, and functions that call it, with a {\tt BruteForce} strategy, this denotes the number of threads to use in brute force point checking.
\end{description}
%Outputs
%:List
%a list of points in the variety with possible repetitions.
%Description
%Text  
%Gives at most $n$ many point in a variety $V(I)$. 
%Example
%R = ZZ/5[t_1..t_3];
%I = ideal(t_1,t_2+t_3);
%randomPoints(3, I)
%randomPoints(4, I, Strategy => Default)
%randomPoints(4, I, Strategy => GenericProjection)
%randomPoints(4, I, Strategy => LinearIntersection)

@@

			i2 : R = ZZ/7[x_1..x_10]

			o2 = R

			o2 : PolynomialRing

			i3 : I = ideal(random(2, R), random(3,R));

@@

			i4 : time randomPoints(I,Strategy=>BruteForce, PointCheckAttempts=>20000)
				 -- used 0.0102075 seconds

@@

			i5 : time randomPoints(I,Strategy=>Default)
				 -- used 0.085741 seconds

@@

			i6 : S = ZZ/211[x_1..x_10]

			o6 = S

			o6 : PolynomialRing

@@

			i8 : time randomPoints(J,Strategy=>BruteForce, PointCheckAttempts=>2000000)
				-- used 25.3479 seconds

@@

			i9 : time randomPoints(J,Strategy=>Default)
			-- used 0.102918 seconds

@@

			i10 : time randomPoints(J,Strategy=>LinearIntersection)
			-- used 0.0169781 seconds

@@

			i11 : time randomPoints(J,Strategy=>GenericProjection)
			-- used 0.364902 seconds

@@

			i12 : time randomPoints(J,Strategy=>LinearIntersection, ExtendField => true)
     		-- used 0.0612306 seconds

@@

	\vspace{1em}
	-- {\tt findANonZeroMinor(n,M,I)}

@@

	~~
	{{\small\color{blue}
		\begin{verbatim}
		i3 : R = ZZ/5[x,y,z]

		o3 = R

		o3 : PolynomialRing

		i4 : I = ideal(random(3,R)-2, random(2,R))

@@

		i6 : findANonZeroMinor(2,M,I, Strategy => GenericProjection)

@@

		o6 : Sequence
		\end{verbatim}
	}}
\end{example}

@@

\vspace{1em}
-- {\tt extendIdealByNonZeroMinor(n,M,I)}

@@

%\vspace{1em}
As before, we can have the following additional options:
\begin{description}
\item[\tt Strategy => Symbol] This specifies which strategy to use when calling {\tt randomPoints}.
\item[\tt Homogeneous => Boolean] This is passed in calls to {\tt randomPoints}.
\item[\tt Verbose => Boolean] This turns on or off verbose output (see \Cref{findANonZeroMinor}).
\item[\tt MinorPointAttempts => ZZ] This controls how many points at which to check the rank of the matrix.
\end{description}

@@

	{{\small\color{blue}
\begin{verbatim}
i2 : R = ZZ/5[x,y,z]

o2 = R

o2 : PolynomialRing

i3 : I = ideal(random(3,R)-2, random(2,R))

@@

i5 : extendIdealByNonZeroMinor(2,M,I, Strategy => LinearIntersection)

@@

o5 : Ideal of R
\end{verbatim}	
}}	
\end{example}

@@

	~~
	{{\small\color{blue}
	\begin{verbatim}
		i2 : T = ZZ/101[x1,x2,x3,x4,x5,x6,x7];

@@

		i7 : elapsedTime(while (i < 10) and dim J > 1 do (
                                i = i+1; 
                                J = extendIdealByNonZeroMinor(4, M, J)));
		-- 0.903328 seconds elapsed

@@

	\end{verbatim}
}}

@@

%\end{enumerate}

>>>>

\documentclass[11pt]{amsart}
\usepackage{calc,amssymb,amsthm,amsmath,fullpage}
% \usepackage{mathtools}
\RequirePackage[dvipsnames,usenames, table]{xcolor}
\usepackage{hyperref}
\usepackage[capitalize]{cleveref}
\hypersetup{
  bookmarks,
  bookmarksdepth=3,
  bookmarksopen,
  bookmarksnumbered,
  pdfstartview=FitH,
  colorlinks,backref,hyperindex,
  linkcolor=Sepia,
  anchorcolor=BurntOrange,
  citecolor=MidnightBlue,
  citecolor=OliveGreen,
  filecolor=BlueViolet,
  menucolor=Yellow,
  urlcolor=OliveGreen
}
\usepackage{xypic}
\usepackage{alltt}
\usepackage{multicol}  
% \usepackage{etex}
\usepackage{xspace}
\usepackage{rotating}
\interfootnotelinepenalty=100000
\theoremstyle{definition}
\newtheorem{example}{Example}[section]

% \usepackage{mabliautoref}
% \usepackage{colonequals}
% \frenchspacing
% \input{kmacros3.sty}
% \usepackage{stmaryrd}
% 
\usepackage{verbatim}
\usepackage{enumerate}
\begin{document}
\title{{RandomRationalPoints} package for \emph{Macaulay2}}
\author{Sankhaneel Bisui}
\email{sbisui@tulane.edu}
\address{Tulane University}
\author{Zhan Jiang}
\email{zoeng@umich.edu}
\address{University of Michigan}
\author{Sarasij Maitra}
\email{sm3vg@virginia.edu}
\address{University of Virginia}
\author{Thai Nguyen}
\email{tnguyen11@tulane.edu}
\address{Tulane University}
\author{Karl Schwede}
\email{schwede@math.utah.edu}
\address{Department of Mathematics, University of Utah, 155 S 1400 E Room 233, Salt Lake City, UT, 84112}
\date{\today}

@@

\keywords{RandomRationalPoints, Macaulay2}
% \thanks{Martinova was supported by a University of Utah Mathematics REU fellowship and by the University of Utah ACCESS program.  
%% Yao was supported by a University of Utah Mathematics REU fellowship.}

@@

% We start {\tt randomCoordinateChange}, a helper function that also might be useful to others.

% \item {\tt Codimension}: It takes input an integer. Various functions need to know the codimension/height of the scheme/ideal it is working with.  Setting this to be an integer will tell the function not to compute the codimension and to use this value instead.  The default value is {\tt null}, in which case the function will compute the codimension.  
%%%   
%%   Codimension is not a function, it is an option.
%%%   
\subsection*{\tt randomCoordinateChange:} This function takes a polynomial ring as an input and outputs the coordinate change map.  
i.e. given a polynomial ring, this will produce a linear automorphism of the ring.  This function checks whether the map is an isomorphism by computing the Jacobian.

@@

  {{\small\color{blue}
\begin{verbatim}
	i3 : R = ZZ/11[x, y, z];

@@

                    ZZ
o4 : RingMap R <--- --[x, y, z]
                    11

	i5 : matrix randomCoordinateChange(R, MaxCoordinatesToReplace => 1)

@@

	i6 : matrix randomCoordinateChange(R, MaxCoordinatesToReplace => 1, Homogeneous => false)

@@

	i7 : matrix randomCoordinateChange(R, MaxCoordinatesToReplace => 1, Replacement => Binomial)

@@

% \begin{enumerate}
\subsection{genericProjection} 
This function finds a random (somewhat, depending on options) generic projection of the ring or ideal.
The typical usages are as follows: 
\vspace{1em}

@@

i1 : R = ZZ/5[x, y, z, w];

i2 : I = ideal(x, y^2, w^3 + x^2);

i3 : genericProjection(2, I)

            ZZ                                         2          2
o3 = (map(R,--[z, w],{- x - 2y - z, - y - 2z}), ideal(z  - z*w - w ))                   
            5

o3 : Sequence
\end{verbatim}}
    }}
\end{example}

@@

i1 : R = ZZ/5[x, y, z, w];

i2 : I = ideal(x, y^2, w^3 + x^2);

i3 : genericProjection(2, R/I)

@@

o3 : Sequence
\end{verbatim}}
    }}
\end{example}

@@

i2 : R = ZZ/5[x, y, z, w];

i3 : I = ideal(x, y^2, w^3 + x^2);

@@

i4 :genericProjection(2, I, Replacement => Binomial)

@@

o4 : Sequence
\end{verbatim}}
    }}
\end{example}

% \vspace{1em} 
\subsection{\tt projectionToHypersurface} This function creates a projection to a hypersurface. The typical usages are as follows:

@@

  {{\small\color{blue}
\begin{verbatim}
	i2 : R = ZZ/5[x, y, z];

	i3 : I = ideal(random(3, R) - 2, random(2, R))

@@

\end{verbatim}}
    }}
\end{example}

% \end{enumerate}
\section{randomPoints}\label{randomPoints}
{\tt randomPoints} is a function to find rational or geometric points in a variety. The typical usages are as follows:

@@

  Notice that the speed, or success, varies depending on the strategy.

@@

			i2 : R = ZZ/101[x, y, z];

			i3 : J = ideal(x^3 + y^2 + 1, z^3 - x^2 - y^2 + 2);

@@

\end{verbatim}}
      }}
  \end{example}

@@

    {{\small\color{blue}
\begin{verbatim}
	i1: R = ZZ/101[x, y, z];

	i2 : J = ideal "x3 + y2 + 1, z3 - x2 - y2 + 2";

@@

	i3 : randomPoints(J, Strategy => BruteForce, PointCheckAttempts => 20000)
		-- used 0.84679 seconds

@@

	o3 : List
\end{verbatim}}
      }}
  \end{example}

@@

  When calling {\tt randomPoints}, and functions that call it, with a {\tt BruteForce} strategy, this denotes the number of threads to use in brute force point checking.
\end{description}
% Outputs
% :List
% a list of points in the variety with possible repetitions.
% Description
% Text  
% Gives at most $n$ many point in a variety $V(I)$. 
% Example
% R = ZZ/5[t_1..t_3];
% I = ideal(t_1,t_2+t_3);
% randomPoints(3, I)
% randomPoints(4, I, Strategy => Default)
% randomPoints(4, I, Strategy => GenericProjection)
% randomPoints(4, I, Strategy => LinearIntersection)

@@

			i2 : R = ZZ/7[x_1..x_10];

			i3 : I = ideal(random(2, R), random(3, R));

@@

			i4 : time randomPoints(I, Strategy => BruteForce, PointCheckAttempts => 20000)
				 -- used 0.0102075 seconds

@@

			i5 : time randomPoints(I, Strategy => Default)
				 -- used 0.085741 seconds

@@

			i6 : S = ZZ/211[x_1..x_10];

@@

			i8 : time randomPoints(J, Strategy => BruteForce, PointCheckAttempts => 2000000)
				-- used 25.3479 seconds

@@

			i9 : time randomPoints(J, Strategy => Default)
			-- used 0.102918 seconds

@@

			i10 : time randomPoints(J, Strategy => LinearIntersection)
			-- used 0.0169781 seconds

@@

			i11 : time randomPoints(J, Strategy => GenericProjection)
			-- used 0.364902 seconds

@@

			i12 : time randomPoints(J, Strategy => LinearIntersection, ExtendField => true)
     		-- used 0.0612306 seconds

@@

\vspace{1em}
-- {\tt findANonZeroMinor(n, M, I)}

@@

  ~~
  {{\small\color{blue}
\begin{verbatim}
		i3 : R = ZZ/5[x, y, z];

		i4 : I = ideal(random(3, R) - 2, random(2, R))

@@

		i6 : findANonZeroMinor(2, M, I, Strategy => GenericProjection)

@@

		o6 : Sequence
\end{verbatim}}
    }}
\end{example}

@@

\vspace{1em}
-- {\tt extendIdealByNonZeroMinor(n, M, I)}

@@

% \vspace{1em}
As before, we can have the following additional options:
\begin{description}
\item[\tt Strategy => Symbol] This specifies which strategy to use when calling {\tt randomPoints}.
\item[\tt Homogeneous => Boolean] This is passed in calls to {\tt randomPoints}.
\item[\tt Verbose => Boolean] This turns on or off verbose output (see \Cref{findANonZeroMinor}).
\item[\tt MinorPointAttempts => ZZ] This controls how many points at which to check the rank of the matrix.
\end{description}

@@

  {{\small\color{blue}
\begin{verbatim}
i2 : R = ZZ/5[x, y, z];

i3 : I = ideal(random(3, R) - 2, random(2, R))

@@

i5 : extendIdealByNonZeroMinor(2, M, I, Strategy => LinearIntersection)

@@

o5 : Ideal of R
\end{verbatim}}	
    }}	
\end{example}

@@

  ~~
  {{\small\color{blue}
\begin{verbatim}
		i2 : T = ZZ/101[x1, x2, x3, x4, x5, x6, x7];

@@

		i7 : elapsedTime(while (i < 10) and dim J > 1 do (
                                i = i + 1; 
                                J = extendIdealByNonZeroMinor(4, M, J)));
		-- 0.903328 seconds elapsed

@@

\end{verbatim}}
    }}

@@

% \end{enumerate}
